### 2019-10-22 Manuscript/manuscript.tex
submission version of manuscript

 Paleogeographic snapshots for the past position of Laurentia reconstructed using this synthesis of the paleomagnetic poles are shown in Figure \ref{fig:Laurentia_reconstructions}. These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's accretionary growth. As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves to the longitudinal position of the block. Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}. In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model.

@@

Seeking to developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types. Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles. Paleomagnetic poles compilations themselves are evolving with better data and improved geochronology. Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data. These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted.}}}
\label{fig:Laurentia_reconstructions}
\end{figure}

@@

 An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}. The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and the researchers need to balance varying constraints. The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. go to moderately high southerly latitudes ca. 1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating. The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes.

There are more published models that show snapshots and publish rotation parameters associated with given time intervals such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, but did not publish parameters for a continuous model. The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude}.

@@

There is strong evidence both in the geological and paleomagnetic record of Laurentia for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of the types of low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones in the Proterozoic relative to the Phanerozoic \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. As an example, consider the record of Laurentia in the late Mesoproterozoic. As can be seen in Table 2, there is a fantastic quantity of high-quality paleomagnetic poles between 1110 and 1070 Ma. These poles constrain rapid motion of Laurentia along

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}. The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}}}
\label{fig:Laurentia_reconstructions}
\end{figure}

@@

{\scriptsize
\begin{landscape}
\input{pole_table_edited.tex}
\end{landscape}
}

>>>>

 Paleogeographic snapshots for the past position of Laurentia reconstructed using this synthesis of the paleomagnetic poles are shown in Figure \ref{fig:Laurentia_reconstructions}. These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's accretionary growth. As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves of the longitudinal position of the block. Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}. In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model. The continuous paleolatitude implied by this model is shown as the simple Laurentia pole interpolation mode in Figure \ref{fig:Laurentia_paleolatitude}.

@@

Developing comprehensive global continuous paleogeographic models is a major challenge given the need to integrate and satisfy diverse geological and paleomagnetic data types. Continually improving constraints related to tectonic setting from improved geologic and geochronologic data need to be carefully integrated with the database of paleomagnetic poles. Paleomagnetic poles compilations themselves are evolving with better data and improved geochronology. Efforts such as this volume are therefore essential to present the state-of-the-art in terms of existing constraints that can be used to evaluate current models and set the stage for future progress in Precambrian paleogeography.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic that are well-constrained by paleomagnetic data. These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} and use this model to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted. All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}}}
\label{fig:Laurentia_reconstructions}
\end{figure}

@@

 An exception to the paucity of published continuous paleogeographic models for the Precambrian is the Neoproterozoic model of \cite{Merdith2017b} which is shown in comparison to the constraints for Laurentia in Figure \ref{fig:Laurentia_paleolatitude}. The extent to which the implied position of Laurentia in \cite{Merdith2017b} is consistent with the compiled paleomagnetic constraints can be visualized in Figure \ref{fig:Laurentia_paleolatitude}. As noted above, the development of such models is challenging and the researchers need to balance varying constraints. The focus here will be on the extent to which this model satisfies the available paleomagnetic poles for Laurentia. The model does not honor the Grenville loop (e.g. go to moderately high southerly latitudes ca. 1000 Ma) which is a striking departure from the paleomagnetic record and standard paleogeographic models. Additionally, the implemented plate motion strays from the younger poles of the Keweenawan Track and does not honor the Franklin LIP pole \cite{Denyszyn2009b} despite its `A' Nordic rating. The Franklin pole is taken to be a key constraint at the Tonian/Cryogenian boundary that provides evidence of the supercontinent Rodinia being equatorial and for the Sturtian glaciation having extended to equatorial latitudes \citep{Macdonald2010a}.

There are more published models that show snapshots and publish rotation parameters associated with given time intervals such as the Rodinia model of \cite{Li2008a} and the Mesoproterozoic model of \cite{Pisarevsky2014b}, but did not publish parameters for a continuous model. The position for Laurentia implied by the Euler poles given for the model snapshots of these studies are shown in Figure \ref{fig:Laurentia_paleolatitude} and can be compared to the compiled record.

@@

There is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of the types of low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones in the Proterozoic relative to the Phanerozoic \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. As an example, consider the record of Laurentia in the late Mesoproterozoic. As can be seen in Table 2, there are many high-quality paleomagnetic poles between 1110 and 1070 Ma. These poles constrain rapid motion of Laurentia leading up to collisional orogenesis as illustrated in Figure \ref{fig:Grenville_reconstructions} providing strong evidence for differential plate motion at the time.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
\caption{\small{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}. The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}}}
\label{fig:Grenville_reconstructions}
\end{figure}

@@

\newpage

\textbf{Table 2}: Compilation of paleomagnetic poles from Laurentia
{\scriptsize
\begin{landscape}
\input{pole_table_edited.tex}
\end{landscape}
}

### 2019-10-22 Manuscript/pole_table_edited.tex
submission version of manuscript

\begin{longtable}{p{1 in}p{1 in}rrrrrrrp{1.5 in}}
\toprule
                       terrane &                                        unit name & rating &  site lon (\textdegree) &  site lat (\textdegree) &  plon (\textdegree) &  plat (\textdegree) &  A$_{95}$ &                    age (Ma)&                                     pole reference \\ \hline
\midrule
\endhead
\midrule
\multicolumn{10}{r}{{Continued on next page}} \\ \hline
\midrule
\endfoot

>>>>

\begin{longtable}{p{1 in}p{1 in}rrrrrrrp{1.5 in}}
\toprule
                       terrane &                                        unit name & Nordic rating &  site lon (\textdegree) &  site lat (\textdegree) &  plon (\textdegree) &  plat (\textdegree) &  A$_{95}$ (\textdegree) &                    age (Ma)&                                     pole reference \\ \hline
\midrule
\endhead
\midrule
\multicolumn{10}{r}{{Continued on next page}} \\ \hline
\midrule
\endfoot
